A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-05-07 paper-hierarchical.tex
Updates from Overleaf

\usepackage{epsfig,color}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}

@@

@@

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline performances. Parameters such as flux, dimensions and intensity distribution of the focused beam and coherence properties are obtained from simple analytical calculations to sophisticated computer simulations using ray tracing and waveoptics techniques. A last generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues related to highly demagnifying sychrotron sources and exploiting the beam coherence. The performance of the beamline is studied for two storages rings: the old ESRF (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux and beam sizes, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. They are discussed with emphasis in benefits and limitations of each one. 
\end{abstract}

@@

Many storage ring based x-ray synchrotron facilities are building or planning upgrades to increase brilliance and coherent flux by one to three orders of magnitude.  The first upgrade of a large facility will be the EBS (Extremely Brilliant Source) \cite{orangebook} at the European Synchrotron Radiation Facility (ESRF). The new storage ring of 150 pm emittance that will be commissioned in 2020 will significantly boost the x-ray coherence of the synchrotron beams. Applications exploiting beam coherence such as x-ray photon correlation spectroscopy, coherent diffraction imaging, propagation-based phase contrast imaging and ptychography will strongly benefit from this update.

Accurate calculation and quantitative evaluation of the parameters related to x-ray optics is of paramount importance for designing, building and exploiting the new beamlines. Every modern beamline follows a procedure of conceptual design plus technical design where detailed simulations are essential. These calculations permit to test the design parameters in a virtual computer environment where advantages and limitations can be studied quantitatively. The selection to the best performant optics configuration requires the detailed simulation of the optics imperfections, that in many cases are the limiting factor of the beamline optical configuration. Deformation in the optical elements due to heat load need to be controlled, therefore optics simulations have to include results from engineering modeling of the thermal deformation associated to mechanical issues usually dome using finite eelement analysis. Traditionally the optics calculations for synchrotron beamlines are done using ray tracing techniques. They model the x-ray beam as a collection of different rays that are incoherent one to another. This technique is still essential for obtaining information on aberrations, flux propagation, monochromator bandwidth, etc. However, the high coherence of the new sources make necessary to complement the ray tracing methods with models including effects of diffraction and scattering from a coherent beam. A fully coherent beam is still a dream for the new generation storage rings. In spite of a high increate of the coherent fraction of the new beams, the coherent fraction at tipically hard x-ray photon energies is still at few percent for the new sources. This makes still important a ray tracing analysis that does not become obsolete, but has to be complemented with an analysis based on wave optics. Because of the non perfect coherence the wave optics analysis becomes non trivial and can be done with different levels of approximation, as discussed in this work. In addition, wave optics simulations are usually expensive from the computer point of view.

For illustrating the applicability of a battery of methods and solutions for beamline optics simulations we have chosen the ESRF beamline ID16A. This beamline provides a high-brilliance beam focused to a few nanometer. It shows characteristics increasingly searched in modern beamlines: extremely high demagnification (sub-micrometer beam) and high coherence. It combines coherent imaging techniques and X-ray fluorescence microscopy to perform quantitative 3D characterization of the morphology and the elemental composition of specimens in their native state.

X-ray beamlines are very particular optical systems. Some implications of exploiting the low wavelengths such as the necessity of using grazing incidence reflectors and the use of perfect crystals as typical hard x-ray monochromators make useless, or at least incomplete, the application of common commercial software packages for the study of synchrotron beamlines. For this reason many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW (more than 40 years old) or SRW (almost 30), for mentioning only codes used in this work. These historical tools together with other new applications have been agglutinated in a user friendly software OASYS. Many of the calculations presented here have been performed directly in OASYS or with script created with the help of OASYS.

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and make the selection of the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describe the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximated values of focal size, flux and coherent fraction. Then accuracy is improved using geometrical methods in Section \ref{level1} that give reasonable results as far as some corrections are considered because of the  mentioned characteristics of high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} also include different models whose complexity increases when the results expected are more and mode demanding. The simulations methods are presented in a hierarchical approach:  they provide more and better results but the price to pay is to deal with more complex and computationally expensive tools.

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF in two scenarios. This beamline was built during the Upgrade Phase I at ESRF in \inred{2011??}. It uses one or several undulators as source. During the period 2011-2018 it worked exploiting the ESRF storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. The replacement of this lattice by a new hybrid multi-bend achromat (HMBA) lattice is being achieved in the ESRF Upgrade Phase II. This new lattice will drastically improve the performances of the ESRF source. The result, the Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circunference and electron energy $E_e$. As as result, the x-ray sources will be much coherent, specially in the horizontal direction. This work shows some quantitative results on how the storage ring upgrade will affect a particular beamline is analyzed in this paper.

ID16A \cite{ID16A} is a bemline that provides a nanometric, highly coherent, pink beam. It is dedicated to nanofocusing applications. \todo{please summarize some use and application}. The beamline usually operates at three fixed photon energies: 11200 eV, 17225 eV and 33600 eV. The straight section in the magnet lattice holds three undulators: an in-vacuum undulator of $\lambda_u=26$~mm period, and two other conventional undulators of 18 and 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

@@

ID16A is a long beamline (the distance from the center of the straight section to the sample position is 185 m). A strong focusing is performed with a Kirkpatrick-Baez (KB) set of mirrors placed very close (a few cm) upstream the sample position (focal plane), which ensures a high demagnification. In order to match the acceptance of the KB in the vertical and horizontal planes a focusing multilayer monochromator (ML) has
been chosen. It i) monochromatises the beam with a large energy bandwidth (typically $\Delta E/E \approx 10^{-2}$) to allow high (pink) flux, 
and ii) creates a secondary secondary source in horizontal plane 40 m downstream from the undulator, where a horizontal slit (VSS, virtual source slit) is placed. A beamline schematic is shown in Figure~\ref{fig:ID16A}. The positions of the different elements are shown in Table \ref{tab:Positions}. This table also shows the extremely high geometrical demagnification ($\sim 10^3$) needed for focusing the beam to the nanometric range.

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (Table \ref{tab:Undulators}). This section is devided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

We first estimate the beam size at different positions at the beamline, in particular at the source plane and at the image plane. A first na{\"{i}}ve estimation of the focal size can be done by considering the source size times the magnification factor $M$. In the paraxial approximation, the magnification factor can be expressed as $M=q/p$, where $p$ is the distance between object and imaging element and $q$ is the distance between imaging element and image. The demagnification is the inverse of the magnification ratio: $M^{-1}=p/q$. Systems where $M>1$ are said to be magnifying, whereas systems with $M<1$ (or $M^{-1}>1$) are said to be demagnifying. The latter will be our case, thus is more convenient to work with demagnification ($M^{-1}>1$) rather than with magnification.

The source size is usually approximated supposing a Gaussian distribution of the electron sizes (or divergence) (Table \ref{tab:eBeam}) and undulator natural source size or divergence. Fo some cases, as the undulator emission, the Gaussian approximation may not be accurate and have some implications that will be discussed later. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the electron beam size and divergence standard deviation of the electron distributions at a centre of the symmetry of the straight section in a plane perpendicular to the main propagation direction. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively.

The undulator natural source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \sqrt{\frac{\lambda L}{2 \pi^2}}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

The beam sizes at different positions at the beamline were estimated in a different way depending if the beam is focused or not. If the beam is focused, its dimensions in a first approximation are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam sizes in a position out of the focus the beam divergence is used. In an ideal {\'{e}}tendue preserving optical system, where the conservation of the Smith-Helmholz invariante applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}), one obtains a focal size 139$\times$5 nm$^2$ (H$\times$V) for ESRF and 10$\times$5 nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The focal spot for ESRF lattice is too large in horizontal for the beamline specification, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8 \inred{$\approx$ 403.8/50=8.076}. Therefore, for ESRF, the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

In general, for grazing incident optics imaging extended sources with large demagnifications ($M^{-1} \gg 100$), the effect of optical aberrations is very important even using a perfect optical surface that focus point-to-point. As a consequence, the results calculated using only the demagnification factor are usually too optimistic. Moreover imperfections in the optical surface (figure errors, slope errors and microroughness) are often the limiting factors of the real reflective optics. Their effect will studied in Sections~\ref{level1} and \ref{level2}.
%The smearing of the focal size because of diffractive effects by the mirror clipping will also be discussed.

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$ eV for both the high-$\beta$ (ESRF) and the EBS straight sections. Values are FWHM. Values in parentheses corresponds to closing the VSS slit to 50$\mu$m (21.3 if approximating by the Gaussian $\sigma$).}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    $FWHM_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    $FWHM_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    $FWHM'_h$ [$\mu$rad]:   & 26.96         & \inred{65.22?}        & 189062   & 15.60   & 37.74   & 109404       \\
    $FWHM'_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons at the sample. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most performant and well maintained codes spread in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated over an aperture if 1 $\times$ 1 mm$^2$ placed at $d$=27.1 m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use ($\max(\Sigma') \times d = 0.73$ mm), with $\Sigma'$ values from in Table~\ref{tab:HandCalculations}. The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and 1.9 10$^{14}$ for ESRF (see \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and specially in peak value. The number of photons per second after the multilayer monochromator is 2.9 10$^{15}$ for EBS and 1.9 10$^{15}$ for ESRF, considering that the energy bandwidth transmitted by a multilater monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing at this point ideal reflectivity (100\%) for the multilayers.

@@

To estimate the geometrical transmitivity of the beamline we have to consider which optical elements crop the beam and remove photons. This can be analysed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h=50~\mu$m in horizontal and fully opened in vertical. For ESRF its transmittivity is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their length to a direction perpendicular to the optical axis. Using mirror length of $L_v$=60 mm and $L_h$=26 mm for VFM and HFM, respectively, and incident angle of 15 mrad, the KB is behaving like an aperture of $900 \times 390 \mu m^2$. We get N.A. values of 6.2$\times$21.6 $mu$rad (aperture size divided by the distance from the mirror to the VSS in H and to the source in V). The transmittivity coefficient is the ratio of the accepted N.A. over the beam divergence, resulting 9.5\% (H) and 0.18\% (V)for ESRF and 16.3\% and 0.18\% for EBS. The overall transmission is the product of H and V ratios, therefore 0.2\% for ESRF and 2.95\% for EBS, therefore 3.8 $10^{12}$ photons/s for ESRF and 8.5 $10^{13}$ photons/s for EBS. 
\todo{check with references $10^9$ at 33.6 keV. I get: 1.6e+15 * 0.00295 * 0.7**3 = 1.62e+12. How closed are the slits? We used the undulator peak value over the whole ML bandwidth, that overestimates the integrated flux.}

>>>>

\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{color,xcolor}
\usepackage{siunitx}
\usepackage[normalem]{ulem}

@@

%-------------------------------------------------------------
%%%%%%%%%%% to ignore the \sout{}, uncomment the line below
%\renewcommand{\sout}[1]{\unskip}
%-------------------------------------------------------------
%%%%%%%%%%% to change color from red to black
%ctrl+f, put 'red' in 'Search for' and 'black' in 'Replace with', then press 'All'
%-------------------------------------------------------------

@@

\begin{abstract}
We consider different approaches to simulate a modern X-ray beamline. Several methodologies with increasing complexity are applied to discuss the relevant parameters that quantify the beamline {\color{green}\sout{performances}performance}. Parameters such as flux, dimensions and intensity distribution of the focused beam and coherence properties are obtained from simple analytical calculations to sophisticated computer simulations using ray tracing and {\color{red}\sout{waveoptics}wave optics} techniques. A {\color{green}\sout{last}latest} generation x-ray nanofocusing beamline for coherent  applications (ID16A at the ESRF) has been chosen  to study in detail the issues related to highly demagnifying  {\color{green}\sout{sychrotron}synchrotron} sources and exploiting the beam coherence. The performance of the beamline is studied for two storage{\color{green}\sout{s}} rings: the old ESRF (emittance 4000 pm) and the new ESRF-EBS (emittance 150 pm). In addition to traditional results in term of flux and beam sizes, an innovative study on the partial coherence properties based on propagation of coherent modes is presented. The different algorithms and methodologies are implemented in the software suite OASYS \cite{codeOASYS}. They are discussed with emphasis beamline {\color{green}\sout{in}placed upon the } benefits and limitations of each. 
\end{abstract}

@@

Many storage ring based x-ray synchrotron facilities are building or planning upgrades to increase brilliance and coherent flux by one to three orders of magnitude.  The first upgrade of a large facility will be the EBS (Extremely Brilliant Source) \cite{orangebook} at the European Synchrotron Radiation Facility (ESRF). The new storage ring of 150 pm emittance that will be commissioned in 2020 will significantly boost the  {\color{green}\sout{x-ray}} coherence of the  {\color{green}\sout{synchrotron} x-ray} beams. Applications exploiting beam coherence such as x-ray photon correlation spectroscopy, coherent diffraction imaging, propagation-based phase contrast imaging and ptychography will strongly benefit from this update.

Accurate calculation and quantitative evaluation of the parameters related to x-ray optics is of {\color{green}\sout{paramount} great} importance for designing, building and exploiting {\color{green}\sout{the}} new beamlines. Every modern beamline follows a procedure of conceptual design plus technical design where detailed simulations are essential. These calculations  {\color{green}\sout{permit to test} allowing testing of} the design parameters in a virtual computer environment where advantages and limitations can be studied quantitatively. {\color{green}\sout{The selection to best performant optics configuration} Selection of the optimal optical layout}  requires the detailed simulation of the optics imperfections, that in many cases are the limiting factor of the beamline optical configuration. {\color{green}\sout{Deformation in} Deformations of} the optical elements due to heat load need to be controlled, therefore optics simulations have to include results from engineering modeling of the thermal deformations {\color{green}\sout{associated to mechanical issues}} usually {\color{red}\sout{dome}done} using finite {\color{red}\sout{e}}element analysis. Traditionally the optics calculations for synchrotron beamlines are done using ray tracing techniques. They model the x-ray beam as a collection of {\color{green}\sout{different}incoherent} rays. {\color{green}\sout{that are incoherent one to another}}  This technique is still essential for obtaining information on aberrations, flux propagation, monochromator bandwidth, etc. However, the high coherence of the new sources {\color{green}\sout{make necessary}can introduce the necessity} to complement the ray tracing methods with models including effects of diffraction and scattering from a coherent beam. A fully coherent {\color{green} hard x-ray}beam is still a dream for the new generation storage rings. In spite of a {\color{green}\sout{high increate of }considerable increase in}  the coherent fraction of the new beams, the coherent fraction at {\color{red}typically\sout{tipically}} hard x-ray photon energies is still a few percent for the new sources.  {\color{green}\sout{This makes still important a } Consequently} ray tracing analysis {\color{green}\sout{that}} does not become obsolete, but has to be complemented with an analysis based on wave optics. Because of the {\color{green}\sout{non perfect} partial} coherence the wave optics analysis becomes non trivial and can be done with different levels of approximation, as discussed in this work. In addition, wave optics simulations are usually computationally expensive.\todo{I noticed that there is very few citations to references in the introduction. This is not so nice, as we don't knowledge any previous work. Maybe one could add a few citations in the introduction.}

To illustrate the applicability of a variety of methods and solutions for beamline optical simulations we have chosen the ESRF beamline ID16A~{\color{red}\cite{ID16A}}. This beamline provides a high-brilliance beam focused to a few nanometers. It displays characteristics increasingly aimed for in modern beamlines: extremely high demagnification (sub-micrometer focused beam size) and high coherence. It combines coherent imaging techniques and X-ray fluorescence microscopy to perform quantitative 3D characterization of the morphology and the elemental composition of specimens in their native state~{\color{red}\cite{XRFID16A}}.

X-ray beamlines are very particular optical systems. Some implications of exploiting the {\color{red}\sout{low}short} wavelengths such as the use of grazing incidence reflectors and the use of perfect crystals as typical hard x-ray monochromators render common commercial software packages ill-adapted for the study of synchrotron beamlines. For this reason many optics tools have been developed by the synchrotron community. Some of them have a long history like SHADOW (more than 40 years old) or SRW (almost 30), to mention only codes used in this work. These historical tools together with other new applications have been collected together in a user friendly software package: OASYS. Many of the calculations presented here have been performed directly in OASYS or with scripts created with the help of OASYS.

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. The simulation methods are presented in a hierarchical approach:  they provide more and better results but the price to pay is to deal with more complex and computationally expensive tools.

@@

This section presents the reader with the main operational parameters for the ID16A beamline at ESRF in two scenarios. This beamline was built during the Upgrade Phase I at ESRF in \inred{2011??}. It uses one or several undulators as source. During the period 2011-2018 it operated exploiting the ESRF storage ring’s double-bend achromat (DBA) lattice with horizontal emittance about 4000 pm. In the second phase of the ESRF Upgrade the original storage ring is being replaced by a new hybrid multi-bend achromat (HMBA) lattice. This new lattice will drastically improve the performances of the ESRF source. The resulting Extreme Brilliant Source (EBS) \cite{orangebook} will increase the brightness and coherent fraction by a factor of more than 100, by strongly reducing the horizontal emittance. This is possible using almost the same ring circumference and electron energy $E_e$. As a result, the x-ray sources will be much coherent, specially in the horizontal direction. In this paper, some quantitative results on how the storage ring upgrade will affect a particular beamline are presented. 
%RC190404: @PhaseI -> from the orange book: On the basis of the Purple Book, the ESRF Council decided in June 2008 to launch the ESRF UP by establishing a two-phase process. The Upgrade Programme Phase I (UP PI) for the period 2009-2015, and the Upgrade Programme Phase II (UP PII) to be prepared in the period 2012-2014, and to be presented to the Council for consideration in 2014 and launch in 2015.
% The  Upgrade  Programme  Phase  I  has  been  on  the  ESFRI  Roadmap  since  its  fi rst publication  in  2006,  and  is  today  heading  towards  its  successful  completion  by  the end of 2015.

ID16A \cite{ID16A} is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. \todo{please summarize some use and application}. The beamline can operate at three fixed photon energies: 11200 eV, 17225 eV and 33600 eV. The straight section in the magnet lattice houses three undulators: an in-vacuum undulator of $\lambda_u=26$~mm period, and two other conventional undulators of 18 and 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

%RC190404: From the ID16D page -> The beamline provides a high-brilliance beam focused down to nanometer size, allowing quantitative 3D characterization of the morphology and the elemental composition of specimens in their native state by combining coherent imaging techniques and X-ray fluorescence microscopy.
% The 185 m long beamline provides nano-focused beams for analytical imaging. The nano-imaging beamline addresses problems in biology, biomedicine and nano-technology using X-ray fluorescence microscopy and nano-tomography. It is optimised for ultimate hard X-ray focusing of a beam with a large energy bandwidth at specific energies (17keV or 33.6 keV). Cryogenic sample preservation is available for 2D and 3D imaging.
% Techniques: coherent imaging; phase-contrast imaging; near-field and far-field ptychography; X-ray fluorescence microscopy;  scanning transmission X-ray microscopy; nanotomography
% http://www.esrf.eu/home/UsersAndScience/Experiments/XNP/ID16A/over.html

@@

ID16A is a long beamline (the distance from the center of the straight section to the sample position is 185 m). Strong focusing is performed using a Kirkpatrick-Baez (KB) mirror system placed a few cm upstream of the sample position (focal plane), which ensures a high demagnification. In order to match the acceptance of the KB in the vertical and horizontal planes a focusing multilayer monochromator (ML) has
been chosen. It i) monochromatises the beam with a large energy bandwidth (typically $\Delta E/E \approx 10^{-2}$) to allow high flux, 
and ii) creates a secondary secondary source in the horizontal plane 40 m downstream of the undulator, where a horizontal slit (VSS, virtual source slit) is placed. A beamline schematic is shown in Figure~\ref{fig:ID16A}. The positions of the different elements are shown in Table \ref{tab:Positions}. This table also shows the extremely high geometrical demagnification ($\sim 10^3$) needed for focusing the beam to the nanometric range.

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (Table \ref{tab:Undulators}). This section is divided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

We first estimate the beam size at different positions in the beamline, in particular at the source plane and at the image plane. A first na{\"{i}}ve estimation of the focal size can be made by considering the source size times the magnification factor $M$. In the paraxial approximation, the magnification factor can be expressed as $M=q/p$, where $p$ is the distance between object and imaging element and $q$ is the distance between imaging element and image. The demagnification is the inverse of the magnification ratio: $M^{-1}=p/q$. Systems where $M>1$ are said to be magnifying, whereas systems with $M<1$ (or $M^{-1}>1$) are said to be demagnifying. The latter will be our case, thus is more convenient to work with demagnification ($M^{-1}>1$) rather than with magnification.

The source size (or divergence) is usually approximated supposing a Gaussian distribution of the electron {\color{green}\sout{sizes}positions} (or divergence) (Table \ref{tab:eBeam}) and {\color{green}\sout{undulator natural}single electron photon} source size or divergence. In some cases, such as for undulator emission, this Gaussian approximation may not be accurate and have some implications that will be discussed later. Let $\sigma_{e(h,v)}$ and $\sigma_{e(h,v)}'$ be the standard deviations of electron beam size and divergence  at the centre of  symmetry of the straight section in a plane perpendicular to the main propagation direction. The subscripts $h$ and $v$ stand for horizontal and vertical, respectively.

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \sqrt{\frac{\lambda L}{2 \pi^2}}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

The beam sizes at different positions in the beamline are estimated in different ways depending upon whether the beam is focused or not. If the beam is focused, its dimensions, to a first approximation, are the photon source size multiplied by the corresponding magnification factor. For predicting the photon beam sizes in a position out of the focus the beam divergence is used. In an ideal {\'{e}}tendue conserving optical system, where the Smith-Helmholz invariant applies, the resulting beam divergence transforms inversely as the beam size: it can be calculated as the product between the beam divergence at the source with the demagnification factor $M^{-1}$.

Using the values for U18.3, $n$=1, and  $E=17225$ eV (Table \ref{tab:Undulators}) and the optical configuration described in Table \ref{tab:Positions}, one obtains a focal size 139$\times$5 nm$^2$ (H$\times$V) for ESRF and 10$\times$5 nm$^2$ for EBS (Table~\ref{tab:HandCalculations}). The horizontal focal spot for the ESRF lattice is larger than the specified target value for the beamline, therefore the VSS could be closed to 50$\mu$m thus reducing the beamsize at the VSS plane by a factor ca. 8 \inred{$\approx$ 403.8/50=8.076}. Therefore, for ESRF, the horizontal size at the sample will be affected by the same factor obtaining roughly 17$\times$5 nm$^2$

In general, for simple elliptical cylinder figured grazing incident optics imaging extended sources with large demagnifications ($M^{-1} \gg 100$), the effect of optical aberrations is very important even using a perfect optical surface. As a consequence, the results calculated using only the demagnification factor are usually optimistic. Moreover imperfections in the optical surface (figure errors, slope errors and microroughness) are often the limiting factors of real reflective optics. Their effects will studied in Sections~\ref{level1} and \ref{level2}.
%Effects of slope errors and diffraction limits can be simply estimated by hand calculations as well...
%The smearing of the focal size because of diffractive effects by the mirror clipping will also be discussed.

\begin{table}\label{tab:HandCalculations}
    \centering
    \caption{Photon beam size and divergence for selected ID16A beamline positions. Values below were obtained considering undulator U18.3 tuned to its first harmonic at the photon energy of $E=17225$ eV for both the high-$\beta$ (ESRF) and the EBS straight sections. Values are FWHM. Values in parentheses correspond to closing the VSS slit to 50$\mu$m (21.3 if approximating by the Gaussian $\sigma$).}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{r|ccc|ccc}
                            &               & \textbf{ESRF (high-$\beta$)} &   &         & \textbf{EBS}  &    \\ 
    Element:                & Undulator     & VSS          & Sample            & Undulator & VVS   & Sample       \\
    Position [m]:           & 0.00          & 40.00        & 185.00            & 0.00    & 40.00   & 185.00       \\
    $FWHM_h$ [$\mu$m]:      & 977.19        & 404.00 (50.0)& 0.139 (0.017)     & 71.26   & 29.46   & 0.0102       \\
    $FWHM_v$ [$\mu$m]:      & 9.60          & 482          & 0.0052            & 10.02   & 486.5   & 0.0054       \\
    \newline
    $FWHM'_h$ [$\mu$rad]:   & 26.96         & \inred{65.22?}        & 189062   & 15.60   & 37.74   & 109404       \\
    $FWHM'_v$ [$\mu$rad]:   & 12.04         & 12.04                 & 22264    & 12.16   & 12.16   & 22489        \\                                                   
    \end{tabular}%
    }
\end{table}

Next, we want to estimate the number of photons at the sample. For that, we need the spectrum at the source. It is not straightforward to get this information analytically but one can calculate the flux and power using different codes. Perhaps the most advanced and well maintained codes used in the synchrotron community are SPECTRA \cite{codeSPECTRA} and SRW \cite{codeSRW}. The XOPPY package in the OASYS suite implements three codes: US \cite{codeUS}, URGENT \cite{codeURGENT} and SRW. Figure~\ref{fig:FluxU18} shows the spectra calculated \todo{indicate which code used} over an aperture of 1 $\times$ 1 mm$^2$ placed at $d$=27.1 m. This aperture opening is set to fully accept the central cone at the resonance peak of the harmonic in use ($\max(\Sigma') \times d = 0.73$ mm), with $\Sigma'$ values from in Table~\ref{tab:HandCalculations}. The peak intensity is about  $2.9\cdot10^{14}$ photons/s/0.1{\%}bw for EBS and 1.9 10$^{14}$ for ESRF (see \ref{fig:FluxU18}). One can appreciate how the emittance plays an important role in both peak width and specially in peak value. The number of photons per second after the multilayer monochromator is 2.9 10$^{15}$ for EBS and 1.9 10$^{15}$ for ESRF, considering that the energy bandwidth transmitted by a multilayer monochromator is $\Delta E/E \approx 10^{-2}$ and the bandwidth in Fig.~\ref{fig:FluxU18} is 10$^{-3}$, and supposing, at this point, ideal reflectivity (100\%) for the multilayers.

@@

To estimate the geometrical {\color{red}\sout{transmitivity}transmission?} of the beamline we have to consider which optical elements crop the beam and remove photons. This can be analysed using their angular acceptance or numerical aperture. The multilayer monochromator accepts the whole beam. We consider a VSS closed to $S_h=50~\mu$m in horizontal and fully opened in vertical. For ESRF its {\color{red}\sout{transmittivity}transmission?} is 1/8 (as discussed before) and 1 for EBS. In a similar way, the KB mirrors will crop the beam to a size that is the projection of their {\color{red}useful optical} length onto a direction perpendicular to the optical axis. Using a mirror length of $L_v$=60 mm and $L_h$=26 mm for VFM and HFM, respectively, and incident angle of 15 mrad, the KB behaves as an aperture of $900 \times 390 \mu m^2$. We get N.A. values of 6.2$\times$21.6 $\mu rad^2$ (aperture size divided by the distance from the mirror to the VSS in H and to the source in V). The {\color{red}\sout{transmittivity}transmission?} coefficient is the ratio of the accepted N.A. over the beam divergence, resulting 9.5\% (H) and 0.18\% (V)for ESRF and 16.3\% and 0.18\% for EBS. The overall transmission is the product of H and V ratios, therefore 0.2\% for ESRF and 2.95\% for EBS, therefore 3.8 $10^{12}$ photons/s for ESRF and 8.5 $10^{13}$ photons/s for EBS. 
\todo{check with references $10^9$ at 33.6 keV. I get: 1.6e+15 * 0.00295 * 0.7**3 = 1.62e+12. How closed are the slits? We used the undulator peak value over the whole ML bandwidth, that overestimates the integrated flux.}
\todo{need to consider ML reflectivity for the 3 reflections, also undulator is not perfect, horizontal slit may be closed to less than $50 \mu m$, also KB mirrors may not be fully illuminated and slope error losses due to initial ML mono are neglected}
